{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2025-01-13 paper.tex
Notes and suggestions

\newcommand{\ident}[1]{{\sffamily #1}}
\newcommand{\srcid}[1]{\textnormal{\sffamily #1}}
\newcommand{\srckvd}[1]{\textnormal{\sffamily\bfseries #1}}

@@

We contribute Denicek, a computational substrate that reduces the complexity of
implementing the aforementioned programming experiences. Denicek represents a program as a series of edits
that construct and transform a document consisting of data and formulas. Denicek provides three
primitive operations on series of edits, application of edits, merging of histories and conflict resolution.
We show that the three operations form the backbone of a direct implementation of a range of
programming experiences.

We discuss the architecture of Denicek, document key design considerations and elaborate
the implementation of the programming experiences listed above. To evaluate the proposed
architecture, we use Denicek as the basis of a simple end-user data exploration environment.
The case study shows that the Denicek computational substrate provides an appealing basis for the
design of rich, interactive end-user programming systems.
\end{abstract}

@@

A computational substrate defines the basic structures using which programs are constructed, how
the progam state is represented and how it is changed \cite{jakubovic-2022-ladder}. The choice of
a substrate affects what programming experiences can be readily supported by a system. For example,
object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while using lists as the basic substrate enabled Lisp to become
a ``language laboratory'' \cite{steele-1993-lisp}.

In principle, any programming experience can be developed on top of any computational
substrate. However, a suitable program representation and operations for working with programs
provided by the substrate can eliminate much of the complexity of implementing interesting
programming experiences. For example, the reflective capabilities of Smalltalk make it easy
to provide rich debugging tools \cite{rauch-2019-babylonian} that are difficult to provide
for C/C++~\cite{kell-2018-unix,kell-2024-debugging}.

@@

What would be the ideal programming substrate for rich, interactive end-user programming systems?
In this paper, we aim to design a programming substrate that makes it easy to develop a range of
programming experiences \cite{myers-2006-eup}:

\begin{itemize}
\item \emph{Local-First Collaboration.} Multiple users should be able to concurrently
  modify a single document and merge their changes, preferrably without requiring a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
  demonstrating the steps of the expected behavior using concrete examples~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
  depends on it are invalidated and, possibly, automatically recomputed \cite{mcdirmid-2013-usable,horowitz-2023-engraft,petricek-2020-live}.
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} --
  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
\end{itemize}

@@

Changing data and formulas is done using the same primitive edit operations to
manipulate the document structure. A user-interface may provide a specialized editor, but still
trigger the pritive edits behind the scenes. However, interacting with elements in the document,
such as a entering text in a textbox, also generates a document edit that can be merged or checked
for conflicts (\S\ref{sec:impl-interaction}).

@@

\begin{itemize}
\item We present the Denicek substrate (\S\ref{sec:system}) and provide a detailed description of its
  document representation, edit operations and the key three operations.
\item We explain how a wide range of end-user programming experiences can be readily implemented using
  the substrate (\S\ref{sec:impl}) in the context of a simple web-based prototype (Webnicek).
\item We document important design decisions and alternatives (\S\ref{sec:discuss}). The analysis
  shows that the desired functionality requires a careful choice among interconnected design options.
\item We develop an end-user data exploration system (Datnicek) on top of Denicek (\S\ref{sec:case})
  and evaluate the extent to which the Denicek substrate enables such development (\S\ref{sec:eval}).
\end{itemize}

@@

\subsubsection*{\circled{C} Saving the Interaction} After adding the speaker from the textbox,
the user opens history view, selects edits that added the speaker and saves those as the
{\small\Verb_add-speaker_} interaction. They attach this as a handler for the {\small\Verb_click_} event of the button.

@@

\keyideabox{\faLightbulbO}{Local-First Collaboration}{Denicek's merging
reapplies edits from another branch on top of the current history (\S\ref{sec:impl-collab}).
Merging is asymmetic, but the order does not matter in the above scenario. The same merging
operation is used when handling user interaction (\S\ref{sec:impl-interaction}).}

@@

\subsection{Selectors, Documents and Edits}
A computational document is a tree, consisting of four kinds of nodes (Fig.~\ref{fig:doc}).
References to document locations can be relative or absolute. Both kinds can appear in a
document (reference nodes); absolute references are used in edits (to denote a target~node).

@@

Renaming a field or wrapping a node require corresponding change to references in the document
(\S\ref{sec:system-ops}). When deleting a field to which there is a reference in the document, Denicek
rejects the edit. A copy edit can also be rejected, because it is ambiguous whether references
referring to the original location should continue referring to the source or should be changed to
point to the target of the copying (a reference cannot refer to two locations).

@@

\subsubsection*{Automatic Reference Update}
There are two situations in which automatic update of references is undesirable. If an edit is
applied to a singular element of a list (reference contains the \ident{Index} selector),
references that refer into the list should be unchanged. Although such edits violate the invariant
that collections are homogeneous, they typically do so only temporarily, for example during
construction of a new list item (checking such edits is discussed in \S\ref{sec:discuss-future}).

Documents can also contain values that can be of multiple different kinds (i.e., a union type).
In such case, references should not be updated when the kind of the value changes. For example,
a formula may be either unevaluated or evaluated. As discussed in \S\ref{sec:impl-eval}, evaluation
involves wrapping the formula node, but this should not affect references to the formula.
To support those cases, edits that normally affect selectors (Fig.~\ref{fig:edits}, below)
have a \emph{reference behavior} field that can be set to disable automatic reference
updating. This annotation is required when the target contains the \emph{Index} selector.

\subsubsection*{No Conditional Edits.}
An important aspect of the design is that the effect an edit has on references inside
the document does not depend on the current value of the document. This makes it possible to
define merging solely in terms of edits, without reference to current document state.
(The effect \ident{Copy} has on the document structure depends on the existing document
structure, but not on its value).

@@

\subsection{Primitive Operations}
\label{sec:system-ops}
Denicek provides three primitive operations. A sequence of edits can be applied to a document, two
sequences of edits can be merged and also checked for conflicts. Denicek identifies edit histories by
a (git-like) hash, computed from the hash of the parent and the current edit. The hash is used to
identify a common shared part of the history during conflict resolution and merging.

\subsubsection*{Applying Edits}
When applying an edit, Denicek locates the target node and transform it according to the edit. If
the edit can affect references in the document (Fig~\ref{fig:edits}, below), Denicek updates
matching references in the document according to the rules shown in Fig.~\ref{fig:updates} provided
that the reference behavior of the edit is not set to disable the updating (this is required when
the target reference contains the \ident{Index} selector). Reference update behavior for
\ident{WrapRecord} and \ident{WrapList} differs in that the updated references as a result of
\ident{WrapList} use the \ident{All} selector. Although \ident{WrapList} specifies the index to
be used for the newly created list item, we assume that the operation introduces a homogeneous
list and the new reference should point to all eventual list items.

@@

The key operation that enables such reconcilliation takes two individual edits that occurred
indpendently, $e_1$ and $e_2$, and produces a sequence of edits $e_2', e_2'', \ldots$ that
can be applied after $e_1$ and have the effect of $e_2$, modified to respect the effects of $e_1$.
There are two aspects of such reconcilliation:

@@

@@

\begin{figure}
\newcommand{\eftablecol}[3]{
\small{\bfseries #1}\;\footnotesize\textit{target}\,\; --\,\; {\footnotesize #2}\\[-0.1em]
\quad \small #3\\[0.3em]
}
\begin{tabular}{|p{27em}|}
\hline
\\[-1em]
\eftablecol{StructureEffect}{Affects fields or structure of the target node.}
  {\ident{RenameField}, \ident{DeleteField}, \ident{WrapRecord}, \ident{WrapList}, \ident{Copy}}
\eftablecol{ValueEffect}{Transforms value, modifies list or adds an additional field.}
  {\ident{Add}, \ident{Append}, \ident{Reorder}, \ident{DeleteItem}, \ident{PrimitiveEdit}}
\eftablecol{TagEffect}{Modifies the tage of the target node.}
  {\ident{UpdateTag}}
\hline
\end{tabular}
\vspace{-0.5em}
\caption{Effects of individual edit operations}
\label{fig:effects}
\vspace{-1em}
\end{figure}

@@

\subsubsection*{Conflict Resolution}
When merging two sequences of edits, $E, E_1$ and $E, E_2$, it is sometimes possible to construct
edit histories $E_1'$ and $E_2'$ such that the results of applying $E, E_1, E_2'$ and
$E, E_2, E_1'$ are the same. However, this is not always the case. Two edits can conflict if
they both modify the same value, if they transfrom the structure of a node in incompatible ways
or if one modifies a node nested inside a node replaced or deleted by the other.
Detected conflicts can be reported to the user, but we also use conflict detection to
implement incremental evaluation (\S\ref{sec:impl-incremental}).

@@

@@

The key claim of this paper is that the Denicek computational substrate
makes it easy to support a range of experiences that make programming more concrete,
collaborative and interactive.

In this section, we describe how Webnicek uses the substrate
to support local-first collaboration (\S\ref{sec:impl-collab}), programming by
demonstration (\S\ref{sec:impl-pbd}, \S\ref{sec:impl-interaction}), incremental recomputation
(\S\ref{sec:impl-eval}, \S\ref{sec:impl-incremental}), schema change control (\S\ref{sec:impl-schema}),
end-user debugging via provenance tracking (\S\ref{sec:impl-provenance})
and concrete programming via managed copy \& paste (\S\ref{sec:impl-copy}).

@@

Recall that merging of edit histories is not symmetric. Document $D$ in Fig.~\ref{fig:merging}
can be obtained either by appending $C'$ (produced by the edit reconcilliation operation)
on top of $A,B$ or by appending $B'$ on top of $A,C$. According to our effect analysis, the two
operations are conflicting. Although $B$ primarily affects the document structure, it also adds
a new field to the record (email), which is a \ident{ValueEffect}, conflicting with
\ident{ValueEffect} of $C$. In this scenario, the conflict can be ignored and the resulting document
is the same in both cases.

@@

@@

Second, to replay edits, Webniceks does not append the recorded edits on top of the current history.
When saving edits, it stores the hash of the history at the time of saving. When replaying,
it appends the recorded edits to the top of the original history (at the time of saving) and merges
this new sequence of edits with the current history. This pushes the recorded edits through all
subsequent edits made by the user. The result can be seen in Fig.~\ref{fig:walkthrough} (E), where
a newly added speaker is transformed from a list item to a table row.

Webnicek also accounts for the case where edits recorded in the document are themselves
transformed (when they are reconcilliated with other edits during merging). In this case, Webnicek
updates the recorded edits (an alternative approach is discussed in \S\ref{sec:discuss}).

@@

@@

Denicek documents can contain formulas inspired by the spreadsheet
paradigm \cite{nardi-1990-spreadsheets}. Formulas can specify richer computations than
what can be expressed using document edits. Formulas do not transform the document and their
results are transient, although their evaluation also leverages the substrate operations,
namely merging of edit histories.

As illustrated in Fig.~\ref{fig:counter}, formulas are represented as document nodes with a
special tag ({\small\Verb_<x-formula>_}). They are recognized by a formula evaluator and
rendered in a special way (\circled{1} and \circled{4}), but they are created using ordinary
edits and the Denicek substrate treates them as stnadard nodes.

To evalaute formulas, the formula evaluator generates edits that turn the {\small\Verb_<x-formula>_}
record into {\small\Verb_<x-evaluated>_} (\circled{2} and \circled{3}), which keeps the previous
formula state in the {\small\Verb_formula_} field and the evaluation result in the {\small\Verb_result_} field.
(Keeping the previous state of the formula is not necessary, but it enables provenance analysis
as discussed in \S\ref{sec:impl-provenance}.) The way edits generated by evaluation are merged
with the document is discussed in the next section.

The Counter App example shown in Fig.~\ref{fig:counter} illustrates the interaction between formulas
and programming by demonstration. To implement a counter, the Increment and Decrement buttons
wrap the current counter value in a formula that adds or subtracts 1.

\keyideabox{\faSuperscript}{Formula Language}{
Webnicek exposes the underlying representation of formulas to the user, but the same representation
can be edited through a user-friendly mechanism such as a block-based editor
\cite{jansen-2019-xlblocks} or a calculation view \cite{sarkar-2018-calc}. The key point
is that Denicek's tree structure makes it easy to embed formulas in document in a uniform
way, edit them and merge them with other changes.}

@@

As illustrated in Fig.~\ref{fig:incremental}, Webnicek supports incremental re-computa\-tion.
In the example, the cost of speaker travel depends on the number of speakers, but the cost
of refreshments depends only on two constants. Edit that adds a speaker only invalidates the
former.

The evaluation mechanism is illustrated in Fig.~\ref{fig:eval}. When the formulas are evaluated,
Webnicek generates \emph{evaluated edits} that are appended to the top of the history.
When subsequent edits are made, they are appended after all non-evaluated edits and the
evaluated edits are pushed through the newly added edits. If the edits conflict (according to
conflict detection discussed in \S\ref{sec:system-ops}), affected evaluated edits and edits that
depend on them are dropped.

@@

@@

@@

\subsubsection*{Merging and Document State}
One specific design choice in Denicek is that the merge operation operates solely on two
sequences of edits, but it does not need access to the current document state (or full edit
histories). The reconcilliation of edits (\S\ref{sec:system-ops}) is defined for two individual
edits $e_1$ and $e_2$.

@@

\subsubsection*{Explicit List Indices}
Denicek does not use implicit numerical indices for lists. When adding an item, the
user (or the system built on top of Denicek) has to explicitly provide an index. This design
is motivated by a typical programming by demonstration scenario where the user adds a
new list item and then modifies it (adding a speaker or a Todo item). When using
explicit indices, the edits following \ident{Append} can easily refer to the newly added
item. (To avoid overwriting existing items when replaying recorded edits, Webnicek replaces
indices of newly added items in recorded interactions in \S\ref{sec:impl-pbd}).

@@

The problem could also be avoided by not letting the user modify a list item after adding it.
(We can require new items to be constructed outside of the list and then added through a
single edit that combines the logic of \ident{Append} and \ident{Copy}.) This is easier to
implement, but it requires unintuitive user interaction and it also does not address all
use cases for modifying list item via index.

\subsubsection*{Ordering List Items and Record Fields}
When list indices are not numerical and supplied explicitly, we need another mechanism for
tracking order of list items. Denicek uses a data structure inspired by Mergeable Replicated
Data Types \cite{kaki-2019-mrdts} and requires specifying the index of a preceding list item
when appending to a list. (If two edits append to the same list independently with the same
preceding item, the order is non-deterministic.)

>>>>

\newcommand{\ident}[1]{{\sffamily #1}}
\newcommand{\srcid}[1]{\textnormal{\sffamily #1}}
\newcommand{\srckvd}[1]{\textnormal{\sffamily\bfseries #1}}
\newcommand{\note}[1]{\textcolor{red}{#1}}

@@

We contribute Denicek, a system that reduces the complexity of
implementing the aforementioned programming experiences. Denicek represents a program as a series of edits
that construct and transform a document consisting of data and formulas. Denicek provides three
primitive operations on series of edits: application of edits, merging of histories, and conflict resolution.
We show how these three operations can be composed to implement a range of
programming experiences.
We call Denicek a \emph{computational substrate} because it serves as a shared platform for such varied programming experiences.

We discuss the architecture of Denicek, document key design considerations and elaborate
the implementation of the programming experiences listed above. To evaluate the proposed
architecture, we use Denicek as the basis of a simple end-user data exploration environment.
The case study shows that the Denicek computational substrate provides an appealing basis for the
design of rich, interactive end-user programming systems.

\note{This is a very promising paper. It's so much simpler and more complete than my stuff. I need to consider some of the different choices you've made. I think I am finally getting what your code execution model is. I hope the comments below are helpful.}
\end{abstract}

@@

A computational substrate defines the basic structures with which programs are constructed, how
the progam state is represented, and how it is changed \cite{jakubovic-2022-ladder}. The choice of
a substrate affects what programming experiences can be readily supported by a system. For example,
object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while representing programs as lists enabled Lisp to become
a ``language laboratory'' \cite{steele-1993-lisp}.

In principle, any programming experience can be developed on top of any computational
substrate. However, a suitable program representation and operations for working with programs
provided by the substrate can reduce the complexity of implementing interesting
programming experiences. For example, the reflective capabilities of Smalltalk make it easy
to provide rich debugging tools \cite{rauch-2019-babylonian} that are difficult to provide
for C/C++~\cite{kell-2018-unix,kell-2024-debugging}.

@@

In this paper, we aim to design a programming substrate that makes it easier to develop a range of
rich, interactive programming experiences \cite{myers-2006-eup}:

\begin{itemize}
\item \emph{Local-First Collaboration.} Multiple users should be able to concurrently
  modify a single document and automatically merge their changes, preferrably without requiring a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
  demonstrating the steps of the expected behavior using concrete examples~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
  depends on it are invalidated and, possibly, automatically recomputed \cite{mcdirmid-2013-usable,horowitz-2023-engraft,petricek-2020-live}.
  \note{cite classic incremental computation literature}
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} --
  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}. \note{find more references}
\end{itemize}

@@

Changing data and formulas is done using the same primitive edit operations to
manipulate the document structure. A user-interface may provide a specialized editor, but still
trigger the primitive edits behind the scenes. However, interacting with elements in the document,
such as a entering text in a textbox, also generates a document edit that can be merged or checked
for conflicts (\S\ref{sec:impl-interaction}).

@@

\begin{itemize}
\item We present the Denicek substrate (\S\ref{sec:system}) and provide a detailed description of its
  document representation, edit operations and the key three operations.
\item We explain how a wide range of end-user programming experiences can be implemented using
  the substrate (\S\ref{sec:impl}) in the context of a simple web-based prototype (Webnicek).
\item We document important design decisions and alternatives (\S\ref{sec:discuss}). The analysis
  shows that the desired functionality requires a careful choice among interconnected design options.
  \note{We also identify problems requiring future work.}
\item We develop an end-user data exploration system (Datnicek) on top of Denicek (\S\ref{sec:case})
  and evaluate the extent to which the Denicek substrate enables such development (\S\ref{sec:eval}).
\end{itemize}

@@

\subsubsection*{\circled{C} Abstracting the Interaction} After adding the speaker from the textbox,
the user opens history view, selects edits that added the speaker and saves those as the
{\small\Verb_add-speaker_} interaction. They attach this as a handler for the {\small\Verb_click_} event of the button.

@@

\keyideabox{\faLightbulbO}{Local-First Collaboration}{Denicek's merging
reapplies edits from another branch on top of the current history (\S\ref{sec:impl-collab}).
Merging is asymmetric, but the order does not matter in the above scenario. The same merging
operation is used when handling user interaction (\S\ref{sec:impl-interaction}).}

@@

\subsection{Selectors, Documents and Edits}
A computational document is a tree, consisting of four kinds of nodes (Fig.~\ref{fig:doc}).
Non-leaf nodes are records or lists. Both of these are named with a \emph{tag} as in HTML. Record fields have \note{unique?} names. Lists contain elements of the same type and assign a unique index to each element \note{It seems that these indexes are meant to be unique IDs, in which case it would be clearer to call them that.}
Leaf nodes can be primitive values (strings and numbers) or references to a location in the document tree.
References can be relative or absolute. Both kinds can appear in a
document (reference nodes); absolute references are used in edits (to denote a target~node).
\note{Why are relative references needed? I only use absolute references.}

@@

Renaming a field or wrapping a node converts any references into that location
(\S\ref{sec:system-ops}). When deleting a field to which there is a reference in the document, Denicek
rejects the edit. A copy edit can also be rejected, because it is ambiguous whether references
referring to the original location should continue referring to the source or should be changed to
point to the target of the copying (a reference cannot refer to two locations).
\note{why not have distinct Copy and Move operations to resolve this?}

@@

\subsubsection*{Automatic Reference Update}
There are two situations in which automatic update of references is undesirable. If an edit is
applied to a singular element of a list (reference contains the \ident{Index} selector),
references that refer into the list should be unchanged. Although such edits violate the invariant
that collections are homogeneous, they typically do so only temporarily, for example during
construction of a new list item (checking such edits is discussed in \S\ref{sec:discuss-future}). \note{Wouldn't this leave dangling references? Wouldn't it be better to evolve the references as these edits occur?} \note{Insertion into a list can default the contents based on the type to avoid such a transient type error.}

Documents can also contain values that can be of multiple different kinds (i.e., a union type).
\note{This is surprsing since there has been no prior mention of union types, and lists are homogeneous.}
In such cases, references should not be updated when the kind of the value changes. For example,
a formula may be either unevaluated or evaluated. As discussed in \S\ref{sec:impl-eval}, evaluation
involves wrapping the formula node, but this should not affect references to the formula.
To support those cases, edits that normally affect selectors (Fig.~\ref{fig:edits})
have a \emph{reference behavior} field that can be set to disable automatic reference
updating. This annotation is required when the target contains the \emph{Index} selector.
\note{I think it would be desireable to avoid all these complications. References should be automatically maintained so they ``just work''.}
\note{I believe that evaluation should not change the type/shape of the document. The space for evaluation results should be be preallocated in some sense, either as extra structure in a formula, or as metadata attached to the tree. That said, I've tried various alternative approaches to this in Subtext without ever settling on a preferred solution.}

\subsubsection*{Edits are ``static''.}
An important principle is that the effect an edit has on references inside
the document does not depend on the current value of the document, only its type. This makes it possible to
define merging solely in terms of edits, without reference to current document state.
(The effect \ident{Copy} has on the document structure depends on the existing document
structure, but not on its value).

@@

\subsection{Primitive Operations}
\label{sec:system-ops}
Denicek provides three primitive operations. A sequence of edits can be applied to a document, two
sequences of edits can be merged and also checked for conflicts. Denicek identifies edit histories by
a (git-like) hash, computed from the hash of the parent and the current edit. The hash is used to
identify a common shared part of the history during conflict resolution and merging.
\note{what exactly does parent mean here? History prefix?}

\subsubsection*{Applying Edits}
When applying an edit, Denicek locates the target node and transforms it according to the edit. If
the edit can affect references in the document (Fig~\ref{fig:edits}, below), Denicek updates
matching references in the document according to the rules shown in Fig.~\ref{fig:updates} provided
that the reference behavior of the edit is not set to disable the updating (this is required when
the target reference contains the \ident{Index} selector). Reference update behavior for
\ident{WrapRecord} and \ident{WrapList} differs in that the updated references as a result of
\ident{WrapList} use the \ident{All} selector. Although \ident{WrapList} specifies the index to
be used for the newly created list item, we assume that the operation introduces a homogeneous
list and the new reference should point to all eventual list items.

@@

\note{It might help to define a function signature for merging to make this clearer: $\mathcal{M}(P, L, R) = M$ or $\mathcal{M}_P(L, R) = M$.}
\note{Merging is often thought of as symmetric, perhaps choose a name that is clearly asymmetric, like rebase, translate, or project.}

The key operation that enables such reconciliation takes two individual edits that occurred
independently, $e_1$ and $e_2$, and produces a sequence of edits $e_2', e_2'', \ldots$ that
can be applied after $e_1$ and have the effect of $e_2$, modified to respect the effects of $e_1$.
There are two aspects of such reconciliation:

@@

\note{I believe that in general a ``backwards'' form of merge is also needed, as in my \emph{retract} and the OT \emph{Exclusion Transform}. A regular merge takes forking edits and composes them. The backwards form converts composed edits into forking edits (and can be combined with a forward merge to recompose in the opposite order). This is needed for cherry picking and reverts, however you don't seem to need that for this paper.}

\note{Related question: how do you merge an Append(target, index, after, node) with DeleteItem(target, after)? A famous counterexample in OT was that two insertions adjacent to a deletion could flip their location if the edits were reordered. I'll dig up the details of this counterexample for you. Some systems handle this by leaving behind a ``tombstone'' on delete to remember the position of the deleted item. I have a novel solution but I'm probably going to go back to tombstones anyway because they are simpler and handy in the diff UI.}

@@

\begin{figure}
\newcommand{\eftablecol}[3]{
\small{\bfseries #1}\;\footnotesize\textit{target}\,\; --\,\; {\footnotesize #2}\\[-0.1em]
\quad \small #3\\[0.3em]
}
\begin{tabular}{|p{27em}|}
\hline
\\[-1em]
\eftablecol{StructureEffect}{Affects fields or structure of the target node.}
  {\ident{RenameField}, \ident{DeleteField}, \ident{WrapRecord}, \ident{WrapList}, \ident{Copy}}
\eftablecol{ValueEffect}{Transforms value, modifies list or adds an additional field.}
  {\ident{Add}, \ident{Append}, \ident{Reorder}, \ident{DeleteItem}, \ident{PrimitiveEdit}}
\eftablecol{TagEffect}{Modifies the tag of the target node.}
  {\ident{UpdateTag}}
\hline
\end{tabular}
\vspace{-0.5em}
\caption{Effects of individual edit operations}
\label{fig:effects}
\vspace{-1em}
\end{figure}

@@

\subsubsection*{Conflict Resolution}
When merging two sequences of edits, $E, E_1$ and $E, E_2$, it is sometimes possible to construct
edit histories $E_1'$ and $E_2'$ such that the results of applying $E, E_1, E_2'$ and
$E, E_2, E_1'$ are the same. However, this is not always the case. Two edits can conflict if
they both modify the same value, if they transfrom the structure of a node in incompatible ways
or if one modifies a node nested inside a node replaced or deleted by the other.
Detected conflicts can be reported to the user, but we also use conflict detection to
implement incremental evaluation (\S\ref{sec:impl-incremental}).
Our treatment of conflicts necessarily differs in this respect from replication algorithms like Operational Transformation~\cite{OT} and Convergent Replicated DataTypes~\cite{CRDT} where conflicts are arbitrarily but deterministically resolved.

@@

\note{The effect-style conflict detection is a nice idea! Removing conflicts before merging is also an interesting idea I hadn't considered. That might allow me to simplify my transformation rules. }

@@

The key claim of this paper is that the Denicek computational substrate
suppports a range of experiences that make programming more concrete,
collaborative, and interactive. In particular, Denicek's ability to merge edits is a unifying mechanism that provides essential features across the range. We demonstrate this claim with the Webnicek web-based programming system, which uses Denicek to support:
\begin{enumerate}
\item Local-first collaboration (\S\ref{sec:impl-collab})
\item Programming By Demonstration (\S\ref{sec:impl-pbd}\S\ref{sec:impl-interaction})
\item Incremental recomputation (\S\ref{sec:impl-eval}, \S\ref{sec:impl-incremental})
\item Schema change control (\S\ref{sec:impl-schema})
\item End-user debugging via provenance tracking (\S\ref{sec:impl-provenance})
\item Concrete programming via managed copy \& paste (\S\ref{sec:impl-copy}).
\end{enumerate}

@@

Recall that merging of edit histories is not symmetric. Document $D$ in Fig.~\ref{fig:merging}
can be obtained either by appending $C'$ (produced by the edit reconciliation operation)
on top of $A,B$ or by appending $B'$ on top of $A,C$. According to our effect analysis, the two
operations are conflicting. Although $B$ primarily affects the document structure, it also adds
a new field to the record (email), which is a \ident{ValueEffect}, conflicting with
\ident{ValueEffect} of $C$. In this scenario, the conflict can be ignored and the resulting document
is the same in both cases.

@@

\note{To guarantee that replicas converge it is necessary to have a global order on merges. OT and CRDT algorithms use timestamps combined with a unique ID for each client to establish this global order without a centralized coordinator. We can do the same. This means you never know if your changes will dominate or whether they will be eventually overriden but people seem to be willing to live with it. Personally, I'm a skeptical: I expect that like with NoSQL people will eventually realize this sucks and go back to centralized systems.}

@@

Second, to replay edits, Webnicek does not append the recorded edits on top of the current history.
When saving edits, it stores the hash of the history at the time of saving. When replaying,
it appends the recorded edits to the top of the original history (at the time of saving) and merges
this new sequence of edits with the current history. This pushes the recorded edits through all
subsequent edits made by the user. The result can be seen in Fig.~\ref{fig:walkthrough} (E), where
a newly added speaker is transformed from a list item to a table row.

Webnicek also accounts for the case where edits recorded in the document are themselves
transformed (when they are reconciliated with other edits during merging). In this case, Webnicek
updates the recorded edits (an alternative approach is discussed in \S\ref{sec:discuss}).

Simply replaying recorded edits provides limited programming capabilities. We discuss in the next section how recorded edits can themselves be edited by the user to generalize and conditionalize their effects. More general and conventional programming constructs are discussed in \S\ref{sec:discuss}.

@@

\note{The above paragraph was surprising to me and I'm not sure I fully understand it even after staring at the figure for a while. Edits are recorded as an AST using x-tags to hide them in the UI? That became clear only after reading the next section. Our audience might be a lot happier if we used a textual grammar. Earlier it was said that edits can't be conditional but then a conditional predicate appears here. This is a special feature of button nodes? Or event handler nodes? This starts to resemble Angular and similar template languages (which I find hard to read).}

@@

Denicek documents can contain formulas inspired by the spreadsheet
paradigm \cite{nardi-1990-spreadsheets}. Formulas can specify richer computations than
what can be expressed using document edits. Formulas do not transform the document and their
results are transient.

Formula evaluation leverages Denicek's ability to merge edit histories.
As illustrated in Fig.~\ref{fig:counter}, formulas are represented as document nodes with a
special tag ({\small\Verb_<x-formula>_}). They are recognized by a formula evaluator and
rendered in a special way (\circled{1} and \circled{4}), but they are created using ordinary
edits and the Denicek substrate treats them as standard nodes.

To evaluate formulas, the formula evaluator generates edits that turn the {\small\Verb_<x-formula>_}
record into {\small\Verb_<x-evaluated>_} (\circled{2} and \circled{3}), which keeps the previous
formula state in the {\small\Verb_formula_} field and the evaluation result in the {\small\Verb_result_} field.
(Keeping the previous state of the formula is not necessary, but it enables provenance analysis
as discussed in \S\ref{sec:impl-provenance}.) The way edits generated by evaluation are merged
with the document is discussed in the next section.

\note{As I've said before, I'd prefer to store the result of the formula inside the <x-formula> record instead of wrapping it on evaluation. This avoids changing the shape of the document on evaluation, which I recall added some complexity earlier.}

The Counter App example shown in Fig.~\ref{fig:counter} illustrates the interaction between formulas
and programming by demonstration. To implement a counter, the Increment and Decrement buttons
wrap the current counter value in a formula that adds or subtracts 1.
\note{I'd like to see the code for the increment button.}

\note{OK maybe I'm finally getting it. You aren't mutating the value of the counter, you're wrapping the value in an increment formula. I never got that in our previous discussions. Presumably that evaluation is now frozen so it serves as an execution trace. This is a lot different from my approach where a formula stays unchanged except for the results. One advantage is that fields contain a record of their computed mutations whereas in my approach that is recorded in the past state of the doc whence the mutation occured.}

\keyideabox{\faSuperscript}{Formula Language}{
Webnicek exposes the underlying representation of formulas to the user, but the same representation
can be edited through a user-friendly mechanism such as a block-based editor
\cite{jansen-2019-xlblocks} or a calculation view \cite{sarkar-2018-calc}. The key point
is that Denicek's tree structure makes it easy to embed formulas in document in a uniform
way, edit them and merge them with other changes. \note{Some readers may disagree that it is easy: AST editing is controversial. }}

@@

As illustrated in Fig.~\ref{fig:incremental}, Webnicek supports incremental recomputation.
In the example, the cost of speaker travel depends on the number of speakers, but the cost
of refreshments depends only on two constants. The edit that adds a speaker invalidates only the former computation.

The evaluation mechanism is illustrated in Fig.~\ref{fig:eval}. \note{This figure is out of order.} When the formulas are evaluated,
Webnicek generates \emph{evaluated edits} that are appended to the top of the history.
\note{What exactly are these evaluated edits?}
When subsequent edits are made, they are appended after all non-evaluated edits and the
evaluated edits are pushed through the newly added edits. If the edits conflict (according to
conflict detection discussed in \S\ref{sec:system-ops}), affected evaluated edits and edits that
depend on them are dropped.

@@

\note{I'm confused. Changing the value of a location conflicts with an edit that evaluates a formula containing a reference to that location? And these evaluation edits are second-class edits that get retroactively dropped to ``unevaluate'' the formula? I'd prefer what you suggest above: formulas are seemingly globally re-evaluated after every change, with incremental computation a hidden perf optimization. }

\note{An important property of incremental computation systems is avoiding redundant execution. Usually this involves a topological sort of the computation dependency graph. We may need to have a story about that to avoid hate from reviewer 2. Personally, I think incremental recomputation is about performance, not the programming experience, so it isn't that important for this paper. }

@@

\note{Nice and simple! My approach tried to be a lot fancier but ended up being too heavywieght. }

@@

\subsubsection*{Merging and Document State}
One specific design choice in Denicek is that the merge operation operates solely on two
sequences of edits, but it does not need access to the current document state (or full edit
histories). The reconciliation of edits (\S\ref{sec:system-ops}) is defined for two individual
edits $e_1$ and $e_2$.

@@

\subsubsection*{Explicit List Indices}
Denicek does not use implicit numerical indices for lists. When adding an item, the
user (or the system built on top of Denicek) has to explicitly provide an index. This design
is motivated by a typical programming by demonstration scenario where the user adds a
new list item and then modifies it (adding a speaker or a Todo item). When using
explicit indices, the edits following \ident{Append} can easily refer to the newly added
item. (To avoid overwriting existing items when replaying recorded edits, Webnicek replaces
indices of newly added items in recorded interactions in \S\ref{sec:impl-pbd}). \note{I don't think that was explained back there.}

@@

\note{I'm still unclear whether your list indices are ordinal numbers like in an array or are unique IDs. Your usage of implicit/explicit isn't obvious to me. The above paragraph seems to imply they are IDs. Note that automatically generating unique IDs in code execution can lead to nondeterminism. If you are using IDs, do they affect equality of lists?}

The problem could also be avoided by not letting the user modify a list item after adding it.
(We can require new items to be constructed outside of the list and then added through a
single edit that combines the logic of \ident{Append} and \ident{Copy}.) This is easier to
implement, but it requires unintuitive user interaction and it also does not address all
use cases for modifying list item via index. \note{I'm still confused, but this sounds like it really isn't a solution afterall.}
